Fusing Filters with Integer Linear Programming

### 2014-06-22 2014betterfusionforfilters/section/05-Benchmarks.tex
typesetting/indentation fixes, etc

We used GLPK as an external solver, which is not particularly fast, but has Haskell bindings.
For these small programs with few combinators, it produces results in under one hundred milliseconds.
For larger programs, with twenty or more combinators to be clustered, GLPK is inadequate.
One randomly generated example with twenty-five combinators took over twenty minutes in GLPK, while the commercial solver CPLEX was able to solve the same example in a second, and another open source solver CBC took one hundred seconds.
For the same example, the unoptimised program with all constraints takes over twenty minutes in CBC, and CPLEX takes one and a half seconds.

@@

% -----------------------------------------------------------------------------
\subsection{Normalize2}
To demonstrate the ILP formulation we will use the @normalize2@ example from \S\ref{s:Introduction}, repeated here:
\begin{code}
  normalize2 :: Array Int -> Array Int
  normalize2 xs
   = let sum1 = fold   (+)  0   xs
         gts  = filter (>   0)  xs
         sum2 = fold   (+)  0   gts
         ys1  = map    (/ sum1) xs
         ys2  = map    (/ sum2) xs
     in (ys1, ys2)
\end{code}
We use the final ILP formulation from \S\ref{s:OptimisedConstraints}. First, we calculate $possible$ -- that is, the nodes which have no fusion-preventing path between them.
\[ \{ \{sum1, gts, sum2\}
 , \{sum1, ys2\}
 , \{gts, sum2, ys1\}
 , \{ys1, ys2\} \} \]

@@

\begin{tabbing}
MMMMM   \= MMMMMMM \= M \= MMMMMMM \= M \= MMMMMMM \= \kill
Minimise   \> $25 \cdot x_{sum1, gts} + 1  \cdot x_{sum1,sum2} + 25 \cdot x_{sum1, ys2} +$ \\
           \> $25 \cdot x_{gts, sum2} + 25 \cdot x_{gts, ys1} + 1 \cdot x_{sum2, ys1} +$ \\
           \> $25 \cdot x_{ys1, ys2}  + 5  \cdot c_{gts} + 5 \cdot c_{ys1} + 5 \cdot c_{ys2} $
\\[1ex]
Subject to 
    \> $-5 \cdot x_{sum1, gts}$  \> $\le$ \> $\pi_{gts} - \pi_{sum1}$  \> $\le$ \> $5 \cdot x_{sum1, gts}$  \\
    \> $-5 \cdot x_{sum1, sum2}$ \> $\le$ \> $\pi_{sum2} - \pi_{sum1}$ \> $\le$ \> $5 \cdot x_{sum1, sum2}$ \\
    \> $-5 \cdot x_{sum1, ys2 }$ \> $\le$ \> $\pi_{ys2 } - \pi_{sum1}$ \> $\le$ \> $5 \cdot x_{sum1, ys2 }$ \\
    \> $-5 \cdot x_{gts,  ys1 }$ \> $\le$ \> $\pi_{ys1 } - \pi_{gts }$ \> $\le$ \> $5 \cdot x_{gts, ys1  }$ \\
    \> $-5 \cdot x_{sum2, ys1 }$ \> $\le$ \> $\pi_{ys1 } - \pi_{sum2}$ \> $\le$ \> $5 \cdot x_{sum2, ys1 }$ \\
    \> $-5 \cdot x_{ys1, ys2  }$ \> $\le$ \> $\pi_{ys2 } - \pi_{ys1 }$ \> $\le$ \> $5 \cdot x_{ys1, ys2  }$ 
\\[1ex]
    \> $   x_{gts, sum2 }$ \> $\le$ \> $\pi_{sum2} - \pi_{gts }$ \> $\le$ \> $5 \cdot x_{gts, sum2 }$ 
\\[1ex]
    \>                     \>       \> $\pi_{sum1} < \pi_{ys1}$ \\
    \>                     \>       \> $\pi_{sum2} < \pi_{ys2}$
\\[1ex]
    \> $ x_{gts,sum2} $    \> $\le$ \> $c_{gts}$
\\[1ex]
    \> $x_{gts, sum2}$     \> $\le$ \> $x_{sum1, sum2}$ \\
    \> $x_{sum1,sum1}$     \> $\le$ \> $x_{sum1, sum2}$ \\
    \> $x_{sum1, gts}$     \> $\le$ \> $x_{sum1, sum2}$
\end{tabbing}
\noindent
One minimal solution to this is:
\begin{tabbing}
MMMMMMMMMMMMMMMMMMMMMMMMMM \= M \= \kill
$x_{sum1, gts},~ x_{sum1, sum1},~ x_{sum1, sum2},~ x_{gts, sum2},~ x_{ys1,  ys2}$
    \> $=$ \> $0$ \\
$x_{sum1, ys2},~ x_{gts, ys1},~   x_{sum2, ys1}$
    \> $=$ \> $1$ 
\\[1ex]
$\pi_{sum1},~ \pi_{gts },~ \pi_{sum2}$
    \> $=$ \> $0$ \\
$\pi_{ys1 },~ \pi_{ys2 }$
    \> $=$ \> $1$ 
\\[1ex]
$c_{gts},~ c_{ys1},~ c_{ys2}$           
    \> $=$ \> $0$
\end{tabbing}
This minimal solution is not unique, though in this case the only other minimal solutions use different $\pi$ values, and denote the same clustering. Looking at just the non-zero variables in the objective function, the value is $25 \cdot x_{sum1,ys2} + 25 \cdot x_{gts,ys1} + 1 \cdot x_{sum2, ys1} = 51$. For illustrative purposes, note that objective function could be reduced by setting $x_{sum1,ys2} = 0$ (fusing $sum1$ and $ys1$), but this conflicts with other constraints. Since $x_{sum1, sum2} = 0$, we require $\pi_{sum1} = \pi_{sum2}$, but also \mbox{$\pi_{sum2} < \pi_{ys2}$}. These constraints cannot be satisfied, so a clustering that fused $sum1$ and $ys2$ would not also permit $sum1$ and $sum2$ to be fused.

We will now compare the clustering produced by our system, with the one implied by stream fusion. As described in \cite{lippmeier2013flow}, stream fusion cannot fuse a produced array into multiple consumers, or fuse operators that are not in a producer-consumer relationship. The corresponding values of the $x_{ij}$ variables are:
\begin{tabbing}
MMMMMMMMMMMMMMMMMMMMMMMMMM \= M \= \kill
$x_{gts, sum2}$
    \> $=$ \> $0$ \\
$x_{sum1, gts}, x_{sum1, sum2}, x_{ys1,  ys2}, x_{sum1, ys2}, x_{gts, ys1 }, x_{sum2, ys1}$
    \> $=$ \> $1$
\end{tabbing}
We can force this clustering to be applied in our integer linear program by adding the above equations as new constraints. Solving the resulting program then yields:
\begin{tabbing}
MMMMMMMMMMMMMMMMMMMMMMMMMM \= M \= \kill
$\pi_{sum1}, \pi_{gts }, \pi_{sum2}$
    \> $=$ \> $0$ \\
$\pi_{ys1 }, \pi_{ys2 }$
    \> $=$ \> $1$ \\
$c_{gts}, c_{ys1}, c_{ys2}$           
    \> $=$ \> $0$
\end{tabbing}
Note that although nodes $sum1$ and $sum2$ have equal $\pi$ values, they are not fused. Conversely, if two nodes have different $\pi$ values then they are never fused.

>>>>

We use GLPK as an external solver, which is not particularly fast, but has readily-available Haskell bindings.
For these small programs with few combinators, it produces results in under one hundred milliseconds.
For larger programs with twenty or more combinators to be clustered, GLPK is inadequate.
One randomly generated example with twenty-five combinators took over twenty minutes in GLPK, while the commercial solver CPLEX was able to solve the same example in a second, and another open source solver CBC took one hundred seconds.
For the same example, the unoptimised program with all constraints took over twenty minutes in CBC, while CPLEX still only took one and a half seconds.

@@

% -----------------------------------------------------------------------------
\subsection{Normalize2}
To demonstrate the ILP formulation we will use the @normalize2@ example from \S\ref{s:Introduction}, repeated here:
\begin{code}
  normalize2 :: Array Int -> Array Int
  normalize2 xs
   = let sum1 = fold   (+)  0   xs
         gts  = filter (>   0)  xs
         sum2 = fold   (+)  0   gts
         ys1  = map    (/ sum1) xs
         ys2  = map    (/ sum2) xs
     in (ys1, ys2)
\end{code}

We use the final ILP formulation from \S\ref{s:OptimisedConstraints}. First, we calculate $possible$ -- that is, the nodes which have no fusion-preventing path between them.
\[ \{ \{sum1, gts, sum2\}
 , \{sum1, ys2\}
 , \{gts, sum2, ys1\}
 , \{ys1, ys2\} \} \]

@@

\begin{tabbing}
MMMMM   \= MMMMMMM \= M \= MMMMMMM \= M \= MMMMMMM \= \kill
Minimise   \> $25 \cdot x_{sum1, gts} + 1  \cdot x_{sum1,sum2} + 25 \cdot x_{sum1, ys2} +$ \\
           \> $25 \cdot x_{gts, sum2} + 25 \cdot x_{gts, ys1} + 1 \cdot x_{sum2, ys1} +$ \\
           \> $25 \cdot x_{ys1, ys2}  + 5  \cdot c_{gts} + 5 \cdot c_{ys1} + 5 \cdot c_{ys2} $
\\[1ex]
Subject to 
    \> $-5 \cdot x_{sum1, gts}$  \> $\le$ \> $\pi_{gts} - \pi_{sum1}$  \> $\le$ \> $5 \cdot x_{sum1, gts}$  \\
    \> $-5 \cdot x_{sum1, sum2}$ \> $\le$ \> $\pi_{sum2} - \pi_{sum1}$ \> $\le$ \> $5 \cdot x_{sum1, sum2}$ \\
    \> $-5 \cdot x_{sum1, ys2 }$ \> $\le$ \> $\pi_{ys2 } - \pi_{sum1}$ \> $\le$ \> $5 \cdot x_{sum1, ys2 }$ \\
    \> $-5 \cdot x_{gts,  ys1 }$ \> $\le$ \> $\pi_{ys1 } - \pi_{gts }$ \> $\le$ \> $5 \cdot x_{gts, ys1  }$ \\
    \> $-5 \cdot x_{sum2, ys1 }$ \> $\le$ \> $\pi_{ys1 } - \pi_{sum2}$ \> $\le$ \> $5 \cdot x_{sum2, ys1 }$ \\
    \> $-5 \cdot x_{ys1, ys2  }$ \> $\le$ \> $\pi_{ys2 } - \pi_{ys1 }$ \> $\le$ \> $5 \cdot x_{ys1, ys2  }$ 
\\[1ex]
    \> $   x_{gts, sum2 }$ \> $\le$ \> $\pi_{sum2} - \pi_{gts }$ \> $\le$ \> $5 \cdot x_{gts, sum2 }$ 
\\[1ex]
    \>                     \>       \> $\pi_{sum1} < \pi_{ys1}$ \\
    \>                     \>       \> $\pi_{sum2} < \pi_{ys2}$
\\[1ex]
    \> $ x_{gts,sum2} $    \> $\le$ \> $c_{gts}$
\\[1ex]
    \> $x_{gts, sum2}$     \> $\le$ \> $x_{sum1, sum2}$ \\
    \> $x_{sum1,sum1}$     \> $\le$ \> $x_{sum1, sum2}$ \\
    \> $x_{sum1, gts}$     \> $\le$ \> $x_{sum1, sum2}$
\end{tabbing}
\noindent

One minimal solution to this is:
\begin{tabbing}
MMMMMMMMMMMMMMMMMMMMMMMMMM \= M \= \kill
$x_{sum1, gts},~ x_{sum1, sum1},~ x_{sum1, sum2},~ x_{gts, sum2},~ x_{ys1,  ys2}$
    \> $=$ \> $0$ \\
$x_{sum1, ys2},~ x_{gts, ys1},~   x_{sum2, ys1}$
    \> $=$ \> $1$ 
\\[1ex]
$\pi_{sum1},~ \pi_{gts },~ \pi_{sum2}$
    \> $=$ \> $0$ \\
$\pi_{ys1 },~ \pi_{ys2 }$
    \> $=$ \> $1$ 
\\[1ex]
$c_{gts},~ c_{ys1},~ c_{ys2}$           
    \> $=$ \> $0$
\end{tabbing}

This minimal solution is not unique, though in this case the only other minimal solutions use different $\pi$ values, and denote the same clustering. Looking at just the non-zero variables in the objective function, the value is $25 \cdot x_{sum1,ys2} + 25 \cdot x_{gts,ys1} + 1 \cdot x_{sum2, ys1} = 51$. For illustrative purposes, note that objective function could be reduced by setting $x_{sum1,ys2} = 0$ (fusing $sum1$ and $ys1$), but this conflicts with other constraints. Since $x_{sum1, sum2} = 0$, we require $\pi_{sum1} = \pi_{sum2}$, but also \mbox{$\pi_{sum2} < \pi_{ys2}$}. These constraints cannot be satisfied, so a clustering that fused $sum1$ and $ys2$ would not also permit $sum1$ and $sum2$ to be fused.

We will now compare the clustering produced by our system, with the one implied by stream fusion. As described in \cite{lippmeier2013flow}, stream fusion cannot fuse a produced array into multiple consumers, or fuse operators that are not in a producer-consumer relationship. The corresponding values of the $x_{ij}$ variables are:
\begin{tabbing}
MMMMMMMMMMMMMMMMMMMMMMMMMM \= M \= \kill
$x_{gts, sum2}$
    \> $=$ \> $0$ \\
$x_{sum1, gts}, x_{sum1, sum2}, x_{ys1,  ys2}, x_{sum1, ys2}, x_{gts, ys1 }, x_{sum2, ys1}$
    \> $=$ \> $1$
\end{tabbing}

We can force this clustering to be applied in our integer linear program by adding the above equations as new constraints. Solving the resulting program then yields:
\begin{tabbing}
MMMMMMMMMMMMMMMMMMMMMMMMMM \= M \= \kill
$\pi_{sum1}, \pi_{gts }, \pi_{sum2}$
    \> $=$ \> $0$ \\
$\pi_{ys1 }, \pi_{ys2 }$
    \> $=$ \> $1$ \\
$c_{gts}, c_{ys1}, c_{ys2}$           
    \> $=$ \> $0$
\end{tabbing}

Note that although nodes $sum1$ and $sum2$ have equal $\pi$ values, they are not fused. Conversely, if two nodes have different $\pi$ values then they are never fused.

### 2014-06-22 2014betterfusionforfilters/section/07-Conclusion.tex
typesetting/indentation fixes, etc

For this work to be useful for Data Parallel Haskell, more combinators are required such as segmented fold and segmented map, which operate on segmented representations of nested arrays.
Rate inference will have to be adapted to handle segmented arrays, possibly using an inner and outer rates.
For example, segmented fold takes a segmented array, and returns a single array of the outer rate.

>>>>

For this work to be useful for Data Parallel Haskell, more combinators are required such as segmented fold and segmented map, which operate on segmented representations of nested arrays.
Rate inference will have to be adapted to handle segmented arrays, possibly using an inner and outer rates.
For example, segmented fold takes a segmented array of inner and outer rate, and returns a single array of the outer rate.
Similarly, segmented append takes two segmented arrays with the same outer rate, but different inner rates, and returns a segmented array with a new inner rate and the same outer rate.
